Image Caption Generation With CLIP+GPT-2 Model

### 2025-05-04 Latex_Paper/Paper.tex
update all

\subsubsection{ Data Preparation}
\begin{equation}
    L = -\sum_{i \in I} \log \frac{\exp(\text{sim}(z_i, z_j)/\tau)}{\sum_{k \neq i}\exp(\text{sim}(z_i, z_k)/\tau)}
    \end{equation}
    where $\text{sim}(z_i, z_j)$ is the cosine similarity between embeddings, and $\tau$ is the temperature hyperparameter.

>>>>

    \subsubsection{Data Preparation}
    After we get the original CLIP embedding, we will normalize it for better training performance. In the actual training process, 
    we found that the trained model may be overly sensitive to some features of some pictures, 
    so we add some noise to 10\% of the feature information of each picture to improve generalization ability and reduce overfitting. 
    These embeddings can then be projected into the GPT-2 space for training.
\pagebreak
    \subsubsection{Model training}
    Next, we use a multi-layer Transformer to map the 
    CLIP embedding to the word embedding space of GPT-2 
    to create a prefix. This prefix is then concatenated 
    with the tokenized caption, which is generated in the data preparation stage,  
    to form the whole input sequence for GPT-2. After receiving these input sequences, 
    the GPT-2 model is trained for a set rounds of training loop to generate the caption of the corresponding image based on the given prefix and token.

    \subsubsection{Optimization method}
    For the weight matrix obtained after each round of training, we use sub-gradient descent 
    to adjust the weights. Additionally, during the entire training process, we use cross-validation to 
    find the optimal solution for multiple hyperparameters in the model, and reasonably control the learning rate to obtain better training results.
\subsection{Loss and objective function for training process}

We use the cross-entropy loss function to measure the difference between the caption generated by the model and the ground-truth captions. The loss function is defined as:

\textbf{The Cross-entropy Loss Function:}
\begin{equation}
\text{Loss} = -\sum_{t=1}^{T} \log P_{\text{GPT-2}}(w_t \mid w_{<t}, \text{prefix})
\end{equation}

Where:  
\begin{itemize}
    \item $T$ — the total number of tokens (words) in the caption  
    \item $w_t$ — the actual word at position $t$ in the ground-truth caption  
    \item $w_{<t}$ — the sequence of all previous words before time step $t$  
    \item $\text{prefix}$ — the mapped image feature vector used to condition the generation  
    \item $P_{\text{GPT-2}}(w_t \mid w_{<t}, \text{prefix})$ — the probability assigned by GPT-2 to word $w_t$ given previous words and the image prefix
\end{itemize}

This loss encourages the model to assign a higher 
probability to the correct next word at each time step. 
Minimizing the cross-entropy loss over the training data 
helps the model generate captions that are more fluent 
and accurate.

\pagebreak
\textbf{The Objective Function in Maximum Likelihood Estimation:}

\begin{equation}
\hat{\theta} = \arg\min_{\theta} -\sum_{i=1}^{N} \sum_{t=1}^{T_i} \log P_\theta(w_t^{(i)} \mid w_{<t}^{(i)}, \text{prefix}^{(i)})
\end{equation}

Where:
\begin{itemize}
    \item $\theta$ — model parameters, including GPT-2 weights and the parameters of the prefix mapping network  
    \item $N$ — the total number of training samples (image-caption pairs)  
    \item $T_i$ — the number of tokens in the $i$-th caption  
    \item $w_t^{(i)}$ — the $t$-th word in the $i$-th ground-truth caption  
    \item $w_{<t}^{(i)}$ — all previous words before position $t$ in the $i$-th caption  
    \item $\text{prefix}^{(i)}$ — the mapped image features for the $i$-th image
\end{itemize}

This objective function aims to find the parameter set $\hat{\theta}$ that minimizes the total negative log-likelihood 
across the training dataset. During training, the model updates $\theta$ using optimization 
techniques such as gradient descent, guided by cross-validation to avoid overfitting and improve generalization.
